---
title: "NBA Stat Projections — Player Performance & Trend Analysis"
author: "Sai"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 5.5)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
```

## Overview

This report analyzes 10 seasons (2008–2017) of NBA player box-score data, sourced from
the public Kaggle dataset *"NBA Players stats since 1950"* (filtered to the 2008–2017
window). It's the R/dplyr/ggplot2 companion to an Excel workbook covering the same
dataset — built to demonstrate the same analysis (position benchmarks, season trends,
player projections) in a code-first tool. Where useful, comments below note which
Excel technique a given dplyr/ggplot2 step replaces.

```{r load-data}
nba <- read.csv("/tmp/clean/nba_clean.csv", stringsAsFactors = FALSE)

# Calculated fields mirrored from the Excel "Raw Data" helper columns:
#   PPG/RPG/APG = PTS, TRB, AST each divided by G   (Excel: IF-wrapped division, avoids #DIV/0!)
#   PrimaryPos  = text before the hyphen in Pos      (Excel: LEFT/FIND on combo listings like "SG-SF")
# TRB and AST in the source file are SEASON TOTALS, not per-game rates -- averaging them
# directly (like PPG) would mix two different scales in the same chart, so RPG/APG exist
# for the same reason the Excel workbook adds RPG/APG helper columns instead of using TRB/AST.
nba <- nba %>%
  mutate(
    PPG = if_else(G > 0, PTS / G, NA_real_),
    RPG = if_else(G > 0, TRB / G, NA_real_),
    APG = if_else(G > 0, AST / G, NA_real_),
    PrimaryPos = sub("-.*", "", Pos)
  )

glimpse(nba)
```

## Position Benchmarks (All Seasons Combined)

The Excel workbook computes this with `COUNTIF` (player count) and `AVERAGEIFS`
(PPG/TRB/AST/FG% averages) grouped by the `PrimaryPos` helper column. In dplyr the
same two-step "filter then aggregate" logic collapses into one `group_by() %>% summarise()`
call — this is the main structural difference between spreadsheet and code approaches:
Excel repeats one formula per row/category, dplyr expresses the grouping once.

```{r position-benchmarks}
position_order <- c("PG", "SG", "SF", "PF", "C")

pos_summary <- nba %>%
  filter(PrimaryPos %in% position_order) %>%
  group_by(PrimaryPos) %>%
  summarise(
    Players = n(),
    Avg_PPG = mean(PPG, na.rm = TRUE),
    Avg_RPG = mean(RPG, na.rm = TRUE),
    Avg_APG = mean(APG, na.rm = TRUE),
    Avg_FG_pct = mean(FG., na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(PrimaryPos = factor(PrimaryPos, levels = position_order)) %>%
  arrange(PrimaryPos)

pos_summary
```

```{r position-chart}
pos_long <- pos_summary %>%
  select(PrimaryPos, Avg_PPG, Avg_RPG, Avg_APG) %>%
  pivot_longer(-PrimaryPos, names_to = "Stat", values_to = "Value")

ggplot(pos_long, aes(x = PrimaryPos, y = Value, fill = Stat)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Avg_PPG" = "#C0392B", "Avg_APG" = "#2980B9", "Avg_RPG" = "#27AE60"),
                     labels = c("Avg_APG" = "Assists", "Avg_PPG" = "Points", "Avg_RPG" = "Rebounds")) +
  labs(title = "Average Points / Rebounds / Assists by Position",
       subtitle = "2008–2017 seasons combined, per-game rates", x = "Position", y = "Per-game average", fill = NULL) +
  theme_minimal(base_size = 13)
```

## Season Trends: League Scoring & Elite Scorers

Mirrors the Excel "Season Rankings" sheet's `AVERAGEIFS` (league avg PPG by season) and
`COUNTIFS` (count of players averaging 25+ PPG per season, a two-condition count: season
AND PPG threshold).

```{r season-trends}
season_trends <- nba %>%
  group_by(Season) %>%
  summarise(
    League_Avg_PPG = mean(PPG, na.rm = TRUE),
    Players_25plus_PPG = sum(PPG >= 25, na.rm = TRUE),
    .groups = "drop"
  )

season_trends
```

```{r season-trend-chart}
ggplot(season_trends, aes(x = Season)) +
  geom_col(aes(y = Players_25plus_PPG), fill = "#F39C12", alpha = 0.6) +
  geom_line(aes(y = League_Avg_PPG * 2), color = "#C0392B", size = 1.2) +
  geom_point(aes(y = League_Avg_PPG * 2), color = "#C0392B", size = 2.5) +
  scale_y_continuous(
    name = "Players averaging 25+ PPG",
    sec.axis = sec_axis(~ . / 2, name = "League avg PPG")
  ) +
  labs(title = "League Scoring Trend vs. Elite Scorer Count, 2008–2017",
       x = "Season") +
  theme_minimal(base_size = 13) +
  theme(axis.title.y.right = element_text(color = "#C0392B"))
```

## Featured Player Trend & Next-Season Projection

The Excel workbook uses `FORECAST.LINEAR` to fit a straight line through each featured
player's 10 seasons of PPG and project 2018. In R, `lm(PPG ~ Season)` is the direct
equivalent — `FORECAST.LINEAR` *is* simple OLS regression under the hood, so this isn't
an approximation, it's the same calculation.

```{r featured-players}
featured <- c("LeBron James", "Kevin Durant", "Dirk Nowitzki", "Dwyane Wade", "Chris Paul")

featured_data <- nba %>%
  filter(Player %in% featured) %>%
  select(Player, Season, PPG) %>%
  arrange(Player, Season)

# One linear model per player, then predict PPG for the season after their last one.
projections <- featured_data %>%
  group_by(Player) %>%
  group_modify(~ {
    fit <- lm(PPG ~ Season, data = .x)
    next_season <- max(.x$Season) + 1
    pred <- predict(fit, newdata = data.frame(Season = next_season))
    tibble(Season = next_season, PPG = as.numeric(pred), Projected = TRUE)
  }) %>%
  ungroup()

featured_data$Projected <- FALSE
featured_full <- bind_rows(featured_data, projections) %>% arrange(Player, Season)

projections %>% select(Player, Projected_Season = Season, Projected_PPG = PPG)
```

```{r featured-chart}
ggplot(featured_full, aes(x = Season, y = PPG, color = Player)) +
  geom_line(data = filter(featured_full, !Projected), size = 1) +
  geom_point(data = filter(featured_full, !Projected), size = 2) +
  geom_point(data = filter(featured_full, Projected), shape = 8, size = 4, stroke = 1.3) +
  labs(title = "PPG Trend & Next-Season Projection, Featured Players",
       subtitle = "Star (*) marks the linear-regression projection for the season after the data ends",
       x = "Season", y = "Points Per Game") +
  theme_minimal(base_size = 13)
```

## Takeaways

Scoring concentrates predictably by position (guards and forwards lead PPG; centers
lead rebounding), the count of 25+ PPG scorers tracks the league-wide pace-up trend
across 2008–2017, and a simple linear trend is a reasonable — though noisy — one-season-out
projection tool for the featured players, consistent with what the Excel `FORECAST.LINEAR`
version of this same workbook shows.

---
*Data source: Kaggle — "NBA Players stats since 1950" (filtered to 2008–2017 seasons).*
