---
title: "Finance & Supply Chain — Profitability and Fulfillment Analysis"
author: "Sai"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 5.5)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
```

## Overview

This report analyzes ~9,000 order-line records from a retail supply-chain dataset
(DataCo-style order/shipment data), covering department, profitability, and on-time
delivery performance. It's the R/dplyr companion to an Excel workbook covering the same
dataset — the department grouping and risk-flag logic below are direct translations of
that workbook's `VLOOKUP` + nested-`IF` helper columns, included here specifically
because this project (unlike the sports/gaming ones) is meant to read as ordinary
day-to-day business analytics, not a hobby dataset.

```{r load-data}
fin <- read.csv("/tmp/clean/finance_clean.csv", stringsAsFactors = FALSE)
glimpse(fin)
```

## Department Grouping (Lookup) & Risk Classification

The raw `Department` field has 11 values that map many-to-one onto 5 broader groups.
Excel does this with `VLOOKUP` against a small reference table; here it's a named R
vector + a join (same lookup, different mechanics). `RiskFlag` mirrors the Excel
workbook's 3-level nested `IF`: a true loss is flagged first regardless of delivery
status; otherwise a late-risk shipment is flagged; otherwise a thin (<10%) but positive
margin is flagged "Low Margin"; everything else is "Healthy."

```{r dept-risk}
dept_groups <- c(
  "Golf" = "Sporting Goods", "Fan Shop" = "Sporting Goods",
  "Apparel" = "Apparel & Footwear", "Footwear" = "Apparel & Footwear",
  "Outdoors" = "Outdoor & Recreation", "Fitness" = "Outdoor & Recreation",
  "Technology" = "Technology",
  "Book Shop" = "Other Specialty", "Discs Shop" = "Other Specialty",
  "Health and Beauty " = "Other Specialty", "Pet Shop" = "Other Specialty"
)

fin <- fin %>%
  mutate(
    DeptGroup = recode(Department, !!!dept_groups),
    RiskFlag = case_when(
      ProfitRatio < 0 ~ "Loss",
      LateDeliveryRisk == 1 ~ "Late Risk",
      ProfitRatio < 0.1 ~ "Low Margin",
      TRUE ~ "Healthy"
    )
  )

fin %>% count(DeptGroup, sort = TRUE)
```

## Company-Wide KPIs

Mirrors the Excel "Profitability Dashboard" KPI strip — same seven headline numbers,
computed with `sum()`/`mean()` instead of `SUM`/`COUNTIF`/`SUMIF` formulas.

```{r kpis}
kpis <- fin %>%
  summarise(
    Total_Sales = sum(Sales, na.rm = TRUE),
    Total_Profit = sum(OrderProfit, na.rm = TRUE),
    Overall_Margin = Total_Profit / Total_Sales,
    Total_Order_Lines = n(),
    Late_Delivery_Rate = mean(LateDeliveryRisk == 1, na.rm = TRUE),
    Loss_Making_Lines = sum(RiskFlag == "Loss"),
    Loss_Amount = sum(OrderProfit[RiskFlag == "Loss"], na.rm = TRUE)
  )

kpis %>%
  pivot_longer(everything(), names_to = "KPI", values_to = "Value") %>%
  mutate(Value = case_when(
    KPI %in% c("Total_Sales", "Total_Profit", "Loss_Amount") ~ dollar(Value),
    KPI %in% c("Overall_Margin", "Late_Delivery_Rate") ~ percent(Value, accuracy = 0.1),
    TRUE ~ comma(Value)
  ))
```

## Risk Flag Mix

```{r risk-mix}
risk_mix <- fin %>%
  count(RiskFlag) %>%
  mutate(Share = n / sum(n)) %>%
  arrange(match(RiskFlag, c("Loss", "Late Risk", "Low Margin", "Healthy")))

risk_mix
```

```{r risk-chart}
ggplot(risk_mix, aes(x = "", y = Share, fill = factor(RiskFlag, levels = c("Healthy", "Low Margin", "Late Risk", "Loss")))) +
  geom_col(width = 1) +
  coord_polar("y") +
  scale_fill_manual(values = c("Healthy" = "#27AE60", "Low Margin" = "#F39C12",
                                "Late Risk" = "#E67E22", "Loss" = "#C0392B"), name = "Risk Flag") +
  labs(title = "Order-Line Risk Flag Mix") +
  theme_void(base_size = 13)
```

## Sales & Margin by Department Group

Mirrors the Excel dashboard's department-group rollup table (`SUMIF` against the
VLOOKUP-derived `DeptGroup` column).

```{r dept-group-rollup}
dept_rollup <- fin %>%
  group_by(DeptGroup) %>%
  summarise(Sales = sum(Sales, na.rm = TRUE), Profit = sum(OrderProfit, na.rm = TRUE), .groups = "drop") %>%
  mutate(Margin = Profit / Sales) %>%
  arrange(desc(Sales))

dept_rollup
```

```{r dept-group-chart}
ggplot(dept_rollup, aes(x = reorder(DeptGroup, Sales), y = Sales)) +
  geom_col(fill = "#1F3864") +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  labs(title = "Sales by Department Group", x = NULL, y = "Sales ($)") +
  theme_minimal(base_size = 13)
```

## Sales by Department Group × Order Status

Mirrors the native pivot table on the Excel "Profitability Dashboard" sheet
(DeptGroup on rows, OrderStatus on columns, sum of Sales as the data field).

```{r pivot-equivalent}
pivot_tbl <- fin %>%
  count(DeptGroup, OrderStatus, wt = Sales, name = "Sales") %>%
  pivot_wider(names_from = OrderStatus, values_from = Sales, values_fill = 0)

pivot_tbl
```

## Fulfillment: Late Delivery by Market & Shipping Mode

Mirrors the Excel "Fulfillment Analysis" sheet's two `COUNTIFS`/`AVERAGEIF` tables
(late rate and avg shipping days, split by Market and by ShippingMode).

```{r fulfillment}
by_market <- fin %>%
  group_by(Market) %>%
  summarise(Orders = n(), Late_Rate = mean(LateDeliveryRisk == 1),
            Avg_Days_Actual = mean(ShippingDaysActual, na.rm = TRUE),
            Avg_Days_Scheduled = mean(ShippingDaysScheduled, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Late_Rate))

by_mode <- fin %>%
  group_by(ShippingMode) %>%
  summarise(Orders = n(), Late_Rate = mean(LateDeliveryRisk == 1),
            Avg_Days_Actual = mean(ShippingDaysActual, na.rm = TRUE),
            Avg_Days_Scheduled = mean(ShippingDaysScheduled, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Late_Rate))

by_market
by_mode
```

```{r fulfillment-chart}
ggplot(by_mode, aes(x = reorder(ShippingMode, Late_Rate), y = Late_Rate)) +
  geom_col(fill = "#922B21") +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  labs(title = "Late Delivery Rate by Shipping Mode", x = NULL, y = "Late Rate") +
  theme_minimal(base_size = 13)
```

## Takeaways

Late delivery risk — not gross loss-making — is the dominant drag on order-line health,
and it concentrates heavily in the slower shipping modes, which is exactly where you'd
expect a fulfillment fix to start paying off. Department-group margins are fairly even,
so the bigger profitability lever here looks operational (shipping mode mix) rather than
product-mix driven.

---
*Data source: a retail supply-chain order/shipment dataset (DataCo-style schema), in the
same spirit as the public "DataCo Supply Chain" Kaggle dataset.*
