---
title: "Returns to Education & the Gender Wage Gap — A Mincer Earnings Regression"
author: "Sai"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 5.5)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
```

## Overview

This report estimates a **Mincer earnings equation** — the workhorse model of labor
economics relating log wages to education and experience — on 526 individuals from the
1976 Current Population Survey (CPS), collected by Henry Farber and distributed as the
`wage1` dataset in Wooldridge's *Introductory Econometrics: A Modern Approach*. It's the
econometrics counterpart to the rest of this portfolio: instead of describing data (the
Excel/R sports and business projects) or querying it (the SQL projects), this project uses
it to make and test statistical inferences — multiple regression, nested model comparisons,
heteroskedasticity testing, and hand-derived robust standard errors — the analytical core
of an economics degree.

The central question: **how much of the raw gender wage gap survives once you control for
human capital (education, experience, tenure) and occupation/industry sorting?**

```{r load}
d <- read.csv("wage1.csv")
cat("Observations:", nrow(d), " | Variables:", ncol(d), "\n")
str(d[, c("wage","lwage","educ","exper","tenure","female","married","nonwhite")])
```

**Data quality note (disclosed up front, not buried in a footnote):** the dataset's own
documentation flags that `exper` (total labor-market experience) and `tenure` (years with
current employer) are self-reported and occasionally inconsistent — `tenure` cannot
logically exceed `exper`, yet `r sum(d$tenure > d$exper)` of `r nrow(d)` observations
(`r round(mean(d$tenure > d$exper)*100,1)`%) violate that. Consistent with how Wooldridge
distributes the data, I keep these rows rather than silently dropping them, and revisit
this in the Limitations section.

## Exploratory Data Analysis

### Wage distribution

```{r eda-wage}
ggplot(d, aes(x = wage)) +
  geom_histogram(bins = 30, fill = "#4f46e5", color = "white") +
  labs(title = "Hourly Wage Is Right-Skewed", subtitle = "Why we model log(wage), not wage",
       x = "Wage ($/hr, 1976)", y = "Count") +
  theme_minimal(base_size = 13)
```

Wages range from \$`r min(d$wage)` to \$`r max(d$wage)`/hr with a mean of
\$`r round(mean(d$wage),2)` but a median of only \$`r round(median(d$wage),2)` — the
right skew typical of earnings data, and the textbook reason to model `log(wage)` instead
of the raw level: it both compresses the skew and lets coefficients be read as approximate
percentage changes.

### Raw gender gap, before any controls

```{r eda-gender}
gap_tbl <- d %>%
  group_by(female) %>%
  summarise(mean_wage = mean(wage), n = n()) %>%
  mutate(group = ifelse(female == 1, "Female", "Male"))
print(gap_tbl)
raw_gap_pct <- (gap_tbl$mean_wage[gap_tbl$female==0] - gap_tbl$mean_wage[gap_tbl$female==1]) /
  gap_tbl$mean_wage[gap_tbl$female==0] * 100
cat(sprintf("Raw gap: %.1f%%\n", raw_gap_pct))
```

Men in this sample average \$`r round(gap_tbl$mean_wage[gap_tbl$female==0],2)`/hr vs.
\$`r round(gap_tbl$mean_wage[gap_tbl$female==1],2)`/hr for women — a raw gap of
**`r round(raw_gap_pct,1)`%**. That number conflates two very different things: women in
this sample may have less education/experience on average, *or* women may be paid less for
the *same* human capital. The regressions below try to separate those two stories.

### Education and experience vs. log wage

```{r eda-scatter}
p1 <- ggplot(d, aes(x = educ, y = lwage)) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "#4f46e5") +
  geom_smooth(method = "lm", se = FALSE, color = "#1e1b4b") +
  labs(title = "Education vs. Log Wage", x = "Years of education", y = "log(wage)") +
  theme_minimal(base_size = 12)

p2 <- ggplot(d, aes(x = exper, y = lwage)) +
  geom_jitter(width = 0.3, alpha = 0.4, color = "#0ea5e9") +
  geom_smooth(method = "loess", se = FALSE, color = "#0c4a6e") +
  labs(title = "Experience vs. Log Wage", subtitle = "Loess curve shows diminishing returns",
       x = "Years of experience", y = "log(wage)") +
  theme_minimal(base_size = 12)

p1
p2
```

The education relationship looks roughly linear; the experience relationship visibly bends
over — wages rise with experience but at a decreasing rate, which is why the models below
include `exper` *and* `exper²`.

### Correlation among continuous predictors

```{r eda-corr}
round(cor(d[, c("educ","exper","tenure","lwage")]), 3)
```

`exper` and `tenure` correlate at 0.50 — expected, since people with more total labor-market
experience also tend to have longer job tenure — but not high enough to cause serious
multicollinearity problems in the models below.

## Model 1: Baseline Mincer Equation

$$\log(wage) = \beta_0 + \beta_1 educ + \beta_2 exper + \beta_3 exper^2 + \beta_4 tenure + u$$

```{r model1}
m1 <- lm(lwage ~ educ + exper + expersq + tenure, data = d)
summary(m1)
```

**Reading the coefficients (semi-log form, so each coefficient ≈ a % change in wage):**

- Each additional year of education is associated with a
  **`r round(coef(m1)["educ"]*100,2)`%** increase in wage, holding experience and tenure
  fixed (p `r ifelse(summary(m1)$coefficients["educ",4]<0.001,"< 0.001",round(summary(m1)$coefficients["educ",4],3))`).
- Experience has a positive but *diminishing* return: the quadratic term is negative
  (`r round(coef(m1)["expersq"],5)`, p `r ifelse(summary(m1)$coefficients["expersq",4]<0.001,"< 0.001",round(summary(m1)$coefficients["expersq",4],3))`),
  meaning the wage gain from one more year of experience shrinks as experience accumulates.
- Solving $-\beta_2 / (2\beta_3) = 0$ for the turning point:

```{r turning-point}
turning_pt <- -coef(m1)["exper"] / (2 * coef(m1)["expersq"])
cat(sprintf("Experience turning point: %.1f years\n", turning_pt))
```

  Predicted wage (holding education/tenure fixed) peaks around
  **`r round(turning_pt,1)` years** of experience and is predicted to *decline* after that —
  a standard, textbook-consistent finding for cross-sectional wage data.
- Tenure adds a further **`r round(coef(m1)["tenure"]*100,2)`%** per year *with the same
  employer*, on top of general experience — consistent with firm-specific human capital /
  internal promotion effects.
- The model explains **`r round(summary(m1)$r.squared*100,1)`%** of the variation in log
  wages (adj. R² = `r round(summary(m1)$adj.r.squared,3)`) — human capital alone leaves most
  of the variation in earnings unexplained, which motivates adding more controls below.

## Model 2: Adding Demographic Controls

```{r model2}
m2 <- lm(lwage ~ educ + exper + expersq + tenure + female + married + nonwhite, data = d)
summary(m2)
```

```{r model2-interp}
female_pct_m2 <- (exp(coef(m2)["female"]) - 1) * 100
```

The headline number: **female** carries a coefficient of `r round(coef(m2)["female"],4)`,
i.e. women earn about **`r round(abs(female_pct_m2),1)`%** less than men with *identical*
education, experience, and tenure (p < 0.001) — almost as large as the unconditional raw
gap. `married` (+`r round(coef(m2)["married"]*100,1)`%) and `nonwhite`
(`r round(coef(m2)["nonwhite"]*100,1)`%) are not statistically distinguishable from zero in
this sample (p = `r round(summary(m2)$coefficients["married",4],3)` and
`r round(summary(m2)$coefficients["nonwhite",4],3)` respectively) — this dataset simply
doesn't have the power, or the gap, to detect those effects.

**Did adding this block of variables matter, jointly?**

```{r anova-1v2}
anova(m1, m2)
```

The F-test rejects the null that female/married/nonwhite jointly add nothing
(p < 0.001) — even though two of the three are individually insignificant, the block is
clearly informative as a whole (driven by `female`).

## Model 3: Adding Occupation & Region

```{r model3}
m3 <- lm(lwage ~ educ + exper + expersq + tenure + female + married + nonwhite +
            profocc + clerocc + servocc + south + west + northcen + smsa, data = d)
summary(m3)
```

```{r model3-interp}
female_pct_m3 <- (exp(coef(m3)["female"]) - 1) * 100
```

This is the most interesting result in the report. If the gender wage gap were mostly a
story of women sorting into lower-paying occupations or regions, controlling for occupation
and region should shrink the `female` coefficient substantially. It barely moves: from
**`r round(abs(female_pct_m2),1)`%** in Model 2 to **`r round(abs(female_pct_m3),1)`%**
here. Occupational and regional sorting explain only a small slice of the gap in this
sample — most of it persists *within* occupation and region.

The occupation/region controls themselves are informative on their own terms: professional
occupations carry a **+`r round(coef(m3)["profocc"]*100,1)`%** premium and service
occupations a **`r round(coef(m3)["servocc"]*100,1)`%** penalty (both p < 0.01) relative to
the baseline occupation category, and living in a metro area (`smsa`) is worth
**+`r round(coef(m3)["smsa"]*100,1)`%** (p < 0.001). Clerical occupation and the regional
dummies (south/west/northcen) are not statistically significant here.

```{r anova-2v3}
anova(m2, m3)
```

The occupation/region block is also jointly significant (p < 0.001) — it improves the
model's fit, just not by reshaping the gender coefficient much.

### Model comparison summary

```{r model-compare-table}
comp <- data.frame(
  Model = c("1: Human capital", "2: + demographics", "3: + occupation/region"),
  R2 = c(summary(m1)$r.squared, summary(m2)$r.squared, summary(m3)$r.squared),
  Adj_R2 = c(summary(m1)$adj.r.squared, summary(m2)$adj.r.squared, summary(m3)$adj.r.squared),
  Female_coef_pct = c(NA, round(female_pct_m2,1), round(female_pct_m3,1))
)
comp$R2 <- round(comp$R2, 3)
comp$Adj_R2 <- round(comp$Adj_R2, 3)
comp
```

```{r gap-chart, fig.height=5}
gap_df <- data.frame(
  stage = factor(c("Raw gap\n(no controls)","+ Human capital\n+ demographics","+ Occupation\n+ region"),
                 levels = c("Raw gap\n(no controls)","+ Human capital\n+ demographics","+ Occupation\n+ region")),
  gap = c(raw_gap_pct, abs(female_pct_m2), abs(female_pct_m3))
)
ggplot(gap_df, aes(x = stage, y = gap)) +
  geom_col(fill = "#4f46e5", width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", gap)), vjust = -0.5, size = 5, fontface = "bold") +
  labs(title = "Gender Wage Gap Barely Narrows After Adding Controls",
       subtitle = "Female wage penalty across model specifications", x = NULL, y = "Wage gap (%)") +
  ylim(0, max(gap_df$gap) * 1.2) +
  theme_minimal(base_size = 13)
```

## Heteroskedasticity: Does It Matter Here?

OLS standard errors assume constant error variance (homoskedasticity). Earnings data
usually violates this in *levels* — high earners have more dispersed pay than low earners.
A Breusch-Pagan test on the squared residuals confirms it:

```{r bp-levels}
m3_level <- lm(wage ~ educ + exper + expersq + tenure + female + married + nonwhite +
                  profocc + clerocc + servocc + south + west + northcen + smsa, data = d)
u_level <- resid(m3_level)
bp_level <- lm(u_level^2 ~ educ + exper + expersq + tenure + female + married + nonwhite +
                  profocc + clerocc + servocc + south + west + northcen + smsa, data = d)
n <- nrow(d); k_bp <- length(coef(bp_level)) - 1
bp_stat_level <- n * summary(bp_level)$r.squared
bp_p_level <- 1 - pchisq(bp_stat_level, df = k_bp)
cat(sprintf("Levels model (wage ~ ...): BP stat = %.2f, df = %d, p = %.2e\n", bp_stat_level, k_bp, bp_p_level))
```

Strongly significant (p `r format(bp_p_level, scientific=TRUE, digits=3)`) — exactly why
labor economists model `log(wage)`, not `wage`, in the first place. Re-running the same
test on the **log** model:

```{r bp-logs}
u3 <- resid(m3)
bp_log <- lm(u3^2 ~ educ + exper + expersq + tenure + female + married + nonwhite +
                  profocc + clerocc + servocc + south + west + northcen + smsa, data = d)
bp_stat_log <- n * summary(bp_log)$r.squared
bp_p_log <- 1 - pchisq(bp_stat_log, df = k_bp)
cat(sprintf("Log model (lwage ~ ...): BP stat = %.2f, df = %d, p = %.3f\n", bp_stat_log, k_bp, bp_p_log))
```

The log transform does most of the work: heteroskedasticity is no longer statistically
significant at conventional levels (p = `r round(bp_p_log,3)`). There's a mild residual
pattern in the tails though:

```{r resid-by-tercile}
ft <- fitted(m3)
grp <- cut(ft, quantile(ft, c(0, 1/3, 2/3, 1)), include.lowest = TRUE, labels = c("Low","Mid","High"))
tapply(u3, grp, sd)
```

Residual spread is noticeably wider in the top fitted-value tercile, so as a robustness
check (best practice regardless of the test result) I compute **heteroskedasticity-robust
(White/HC1) standard errors by hand** — no `sandwich`/`lmtest` package, just the sandwich
formula directly:

$$\widehat{Var}_{HC1}(\hat\beta) = (X'X)^{-1} X' \,\text{diag}(\hat u_i^2)\, X (X'X)^{-1} \cdot \frac{n}{n-k}$$

```{r robust-se}
X <- model.matrix(m3)
u <- resid(m3)
n <- nrow(X); k <- ncol(X)
XtX_inv <- solve(t(X) %*% X)
meat <- t(X) %*% diag(u^2) %*% X * (n / (n - k))
vcov_hc1 <- XtX_inv %*% meat %*% XtX_inv
se_robust <- sqrt(diag(vcov_hc1))
se_classical <- sqrt(diag(vcov(m3)))

se_compare <- data.frame(
  Coefficient = names(coef(m3)),
  Estimate = round(coef(m3), 4),
  SE_classical = round(se_classical, 4),
  SE_robust = round(se_robust, 4),
  p_robust = round(2 * pt(-abs(coef(m3) / se_robust), df = n - k), 4)
)
se_compare
```

The robust and classical standard errors are close for almost every coefficient —
`tenure`'s SE rises the most (`r round(se_classical["tenure"],4)` → `r round(se_robust["tenure"],4)`,
about `r round(se_robust["tenure"]/se_classical["tenure"],2)`x) — but no
significance conclusion at the 5% level flips either direction. The mild heteroskedasticity
detected above doesn't change the substantive findings; it's reported because checking is
the correct intermediate-econometrics move, not because it overturns anything.

## Residual Diagnostics

```{r diagnostics}
par(mfrow = c(1, 2))
plot(fitted(m3), resid(m3), pch = 16, col = rgb(0.31, 0.27, 0.9, 0.4),
     main = "Residuals vs. Fitted", xlab = "Fitted log(wage)", ylab = "Residual")
abline(h = 0, col = "red", lty = 2)
qqnorm(resid(m3), pch = 16, col = rgb(0.31, 0.27, 0.9, 0.4), main = "Normal Q-Q")
qqline(resid(m3), col = "red")
```

Residuals are centered on zero with a slight funnel/widening at higher fitted values
(consistent with the tercile check above) and roughly normal in the middle of the
distribution with some departure in the tails — typical for wage data, and not severe
enough to invalidate the conclusions, especially since the robust SEs confirm the same
inferences.

## Limitations

This is intermediate applied econometrics, not a causal study, and I want to be specific
about why:

**Omitted ability bias.** The biggest threat to interpreting `educ`'s coefficient as a
*causal* return to schooling is unobserved ability: people with more innate ability or
motivation likely get more education *and* earn more for reasons unrelated to schooling
itself, which biases the education coefficient upward. The standard fix — instrumenting
education with something correlated with schooling but not ability (e.g., compulsory
schooling laws, distance to college) — isn't implemented here; this report estimates an
association, not a causal effect.

**External validity.** This is 1976 U.S. CPS data. Wage *levels* are obviously irrelevant
today, and the labor market has changed enough (composition of occupations, female labor
force participation, returns to specific majors) that even the relative patterns shouldn't
be assumed to hold in 2026. The methodology generalizes; the numbers don't.

**Self-reported, possibly mismeasured data.** As flagged at the top, `r sum(d$tenure > d$exper)`
of `r nrow(d)` observations have `tenure > exper`, a logical impossibility, indicating
measurement error in at least one of those two variables for a small share of the sample.
Measurement error in regressors generally biases coefficients toward zero (attenuation
bias) rather than away from it.

**No interaction effects tested.** A natural extension — does the return to education
differ by gender, or does the experience penalty/premium differ by occupation — would
require interaction terms not estimated here.

## Conclusion

Three nested regressions on 1976 CPS data show: education carries a statistically robust
return (~`r round(coef(m1)["educ"]*100,1)`% per year in the simplest specification),
experience has diminishing returns peaking around `r round(turning_pt,0)` years, and a
gender wage gap of roughly `r round(abs(female_pct_m2),0)`–`r round(abs(female_pct_m3),0)`%
persists even after controlling for education, experience, tenure, marital status,
occupation, and region — meaning the gap is not primarily explained by women and men
sorting into different jobs or having different qualifications, at least along the
dimensions this dataset measures. Heteroskedasticity-robust standard errors were derived by
hand and confirm the findings aren't an artifact of a violated OLS assumption. The honest
caveat is that none of this is causal without an identification strategy for education and
without richer controls for productivity-relevant traits this survey didn't collect.
