---
title: "NFL Betting Trends — Against-the-Spread Performance Analysis"
author: "Sai"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 5.5)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Overview

This report analyzes NFL betting lines and final scores (Kaggle: *"NFL scores and betting
data"*), reconstructing each game's against-the-spread (ATS) and over/under result from
the raw spread/total lines and the actual final score. It's the R/dplyr companion to an
Excel workbook covering the same dataset and the same derived columns — the logic below
is a direct translation of that workbook's `VLOOKUP` + nested-`IF` helper-column chain.

```{r load-data}
nfl <- read.csv("/tmp/clean/nfl_clean.csv", stringsAsFactors = FALSE)
glimpse(nfl)
```

## Team Name → Abbreviation Lookup

`FavoriteTeamID` uses team abbreviations; `HomeTeam`/`AwayTeam` use full names. There's
no shared key, so a lookup table is required either way. The Excel version uses
`VLOOKUP` against a "Team Reference" sheet; here the same table is a named R vector
used inside `recode()` — same join, different syntax. Note two franchises (Rams,
Chargers) map old *and* new city names to one current abbreviation, because the betting
data's `FavoriteTeamID` already uses the current code even in old-city seasons —
the same quirk documented in the Excel workbook's lookup-table comment.

```{r team-lookup}
team_ref <- c(
  "Arizona Cardinals" = "ARI", "Atlanta Falcons" = "ATL", "Baltimore Ravens" = "BAL",
  "Buffalo Bills" = "BUF", "Carolina Panthers" = "CAR", "Chicago Bears" = "CHI",
  "Cincinnati Bengals" = "CIN", "Cleveland Browns" = "CLE", "Dallas Cowboys" = "DAL",
  "Denver Broncos" = "DEN", "Detroit Lions" = "DET", "Green Bay Packers" = "GB",
  "Houston Texans" = "HOU", "Indianapolis Colts" = "IND", "Jacksonville Jaguars" = "JAX",
  "Kansas City Chiefs" = "KC", "Los Angeles Chargers" = "LAC", "Los Angeles Rams" = "LAR",
  "Miami Dolphins" = "MIA", "Minnesota Vikings" = "MIN", "New England Patriots" = "NE",
  "New Orleans Saints" = "NO", "New York Giants" = "NYG", "New York Jets" = "NYJ",
  "Oakland Raiders" = "OAK", "Philadelphia Eagles" = "PHI", "Pittsburgh Steelers" = "PIT",
  "San Diego Chargers" = "LAC", "San Francisco 49ers" = "SF", "Seattle Seahawks" = "SEA",
  "St. Louis Rams" = "LAR", "Tampa Bay Buccaneers" = "TB", "Tennessee Titans" = "TEN",
  "Washington Redskins" = "WAS"
)
```

## Deriving Game Results & ATS Outcome

Same five-step chain as the Excel "Raw Data" helper columns:
`HomeAbbrev/AwayAbbrev` (lookup) → `FavoriteSide` (is the favorite home or away?) →
`FavoriteMargin` (the favorite's actual scoring margin, signed) → `ATSResult` (did the
favorite's margin beat the spread?) → `OUResult` (total points vs. the over/under line).
`case_when()` is dplyr's nested-`IF` equivalent — each clause is checked top to bottom,
first match wins, exactly like Excel's nested `IF()`.

```{r derive-results}
nfl <- nfl %>%
  mutate(
    HomeAbbrev = recode(HomeTeam, !!!team_ref),
    AwayAbbrev = recode(AwayTeam, !!!team_ref),
    Winner = case_when(
      HomeScore > AwayScore ~ HomeAbbrev,
      AwayScore > HomeScore ~ AwayAbbrev,
      TRUE ~ "Tie"
    ),
    HomeMargin = HomeScore - AwayScore,
    TotalPoints = HomeScore + AwayScore,
    FavoriteSide = case_when(
      FavoriteTeamID == HomeAbbrev ~ "Home",
      FavoriteTeamID == AwayAbbrev ~ "Away",
      FavoriteTeamID == "PICK" ~ "Pick",
      TRUE ~ "N/A"
    ),
    FavoriteMargin = case_when(
      FavoriteSide == "Home" ~ as.double(HomeMargin),
      FavoriteSide == "Away" ~ as.double(-HomeMargin),
      TRUE ~ NA_real_
    ),
    ATSResult = case_when(
      FavoriteSide == "Pick" ~ "No Line",
      is.na(FavoriteMargin) ~ "N/A",
      (FavoriteMargin + SpreadFavorite) > 0 ~ "Favorite Covered",
      (FavoriteMargin + SpreadFavorite) < 0 ~ "Underdog Covered",
      TRUE ~ "Push"
    ),
    OUResult = case_when(
      TotalPoints > OverUnderLine ~ "Over",
      TotalPoints < OverUnderLine ~ "Under",
      TRUE ~ "Push"
    )
  )

nfl %>% select(Season, HomeTeam, AwayTeam, FavoriteSide, ATSResult, OUResult) %>% head(8)
```

## Season-Level ATS & Over/Under Rates

Mirrors the Excel "League Trends" sheet's `COUNTIFS` (favorite-covers count and
decided-games count, season by season) — `group_by(Season) %>% summarise()` does both
counts in one pass instead of one `COUNTIFS` formula per outcome per season.

```{r season-rates}
season_summary <- nfl %>%
  group_by(Season) %>%
  summarise(
    Games = n(),
    Favorite_Covers = sum(ATSResult == "Favorite Covered"),
    Underdog_Covers = sum(ATSResult == "Underdog Covered"),
    Decided_ATS_Games = Favorite_Covers + Underdog_Covers,
    Favorite_Cover_Rate = if_else(Decided_ATS_Games == 0, NA_real_, Favorite_Covers / Decided_ATS_Games),
    Over_Count = sum(OUResult == "Over"),
    Decided_OU_Games = sum(OUResult %in% c("Over", "Under")),
    Over_Rate = if_else(Decided_OU_Games == 0, NA_real_, Over_Count / Decided_OU_Games),
    .groups = "drop"
  )

season_summary
```

```{r season-rate-chart}
ggplot(season_summary, aes(x = Season)) +
  geom_col(aes(y = Favorite_Cover_Rate), fill = "#7B241C", alpha = 0.85) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Favorite ATS Cover Rate by Season",
       subtitle = "Dashed line = break-even (50%); bookmakers price spreads to land near this line",
       x = "Season", y = "Favorite Cover Rate") +
  theme_minimal(base_size = 13)
```

## Per-Team ATS Record (Home + Away Combined)

Mirrors the Excel "Team ATS Records" sheet's three-criteria `COUNTIFS` (team abbreviation
+ home/away side + ATS result), summed across both home and away appearances for each team.

```{r team-records}
team_records <- bind_rows(
  nfl %>% filter(FavoriteSide == "Home") %>% transmute(Team = HomeAbbrev, ATSResult),
  nfl %>% filter(FavoriteSide == "Away") %>% transmute(Team = AwayAbbrev, ATSResult)
) %>%
  group_by(Team) %>%
  summarise(
    Times_Favored = n(),
    Covers = sum(ATSResult == "Favorite Covered"),
    Cover_Rate = Covers / Times_Favored,
    .groups = "drop"
  ) %>%
  filter(Times_Favored >= 20) %>%
  arrange(desc(Cover_Rate))

team_records
```

```{r team-chart}
top_bottom <- bind_rows(slice_head(team_records, n = 8), slice_tail(team_records, n = 8)) %>%
  mutate(Team = factor(Team, levels = Team))

ggplot(top_bottom, aes(x = Team, y = Cover_Rate, fill = Cover_Rate > 0.5)) +
  geom_col() +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray30") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("TRUE" = "#27AE60", "FALSE" = "#C0392B"), guide = "none") +
  labs(title = "ATS Cover Rate When Favored: Best & Worst 8 Teams (min. 20 games favored)",
       x = NULL, y = "Cover Rate When Favored") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Spread Size vs. Outcome

```{r spread-distribution}
nfl %>%
  filter(!is.na(SpreadFavorite)) %>%
  mutate(SpreadBucket = cut(abs(SpreadFavorite), breaks = c(0, 3, 6, 9, 14, Inf),
                             labels = c("0-3", "3-6", "6-9", "9-14", "14+"))) %>%
  filter(ATSResult %in% c("Favorite Covered", "Underdog Covered")) %>%
  group_by(SpreadBucket) %>%
  summarise(Games = n(), Favorite_Cover_Rate = mean(ATSResult == "Favorite Covered"), .groups = "drop") %>%
  ggplot(aes(x = SpreadBucket, y = Favorite_Cover_Rate)) +
  geom_col(fill = "#2980B9") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray30") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Favorite Cover Rate by Spread Size",
       subtitle = "Does covering get easier/harder as the line gets bigger?",
       x = "Point spread (absolute value)", y = "Favorite Cover Rate") +
  theme_minimal(base_size = 13)
```

## Takeaways

Favorite cover rates cluster close to 50% across seasons, which is exactly what an
efficiently-priced betting market predicts — Vegas sets the line to split action, not
to be "beatable." The per-team and per-spread-bucket splits show some dispersion, but
none of the deviations here are large enough (or based on enough games) to read as a
reliable, repeatable edge.

---
*Data source: Kaggle — "NFL scores and betting data."*
